---
title: "Homework #4: Auto Insurance -- Crash Probability and Cost"
subtitle: "DATA 621 -- Business Analytics and Data Mining"
author:
  - name: Kevin Havis
  - name: Michael Drake
  - name: Rebecca Bronsteain
  - name: Aaliyah John-Henry
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(corrplot)
library(psych)
library(car)
library(caret)
library(pROC)

theme_set(theme_minimal())
```

# Introduction

In this assignment we build two families of models on auto insurance customer data: (1) binary
logistic regression models predicting `TARGET_FLAG`, whether a customer will crash their car, and
(2) multiple linear regression models predicting `TARGET_AMT`, the cost of the crash given that
one occurs.

# Data Exploration

```{r load-data}
#| label: load-data

train <- read_csv("data/insurance-training-data.csv")
evaluation <- read_csv("data/insurance-evaluation-data.csv")

train <- train %>% select(-INDEX)
```

## Size and Structure

```{r data-structure}
#| label: data-structure

# TODO: dim(), glimpse(), etc.
```

## Summary Statistics

```{r summary-stats}
#| label: summary-stats

# TODO: mean / sd / median summary table
```

## Distributions

```{r distributions}
#| label: fig-distributions
#| fig-cap: "Distributions of numeric variables"

# TODO: histograms or density plots of numeric variables
```

## Categorical Variables

```{r categorical}
#| label: fig-categorical
#| fig-cap: "Bar charts of categorical variables"

# TODO: bar charts of categorical variables
```

## Correlation with Targets

```{r correlation}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of numeric variables"

# TODO: correlation matrix / correlation with TARGET_FLAG and TARGET_AMT
```

## Missing Data

```{r missing-data}
#| label: missing-data

# TODO: count of missing values per variable
```

_Discuss what the exploration reveals: notable distributions, skew, outliers, class balance of
`TARGET_FLAG`, the zero-inflated nature of `TARGET_AMT`, correlations, and variables needing
imputation or cleanup (e.g., currency-formatted strings like `INCOME`, `HOME_VAL`, `BLUEBOOK`,
`OLDCLAIM`)._

# Data Preparation

## Cleaning Currency and Categorical Fields

```{r clean-fields}
#| label: clean-fields

# TODO: strip $ and , from currency fields (INCOME, HOME_VAL, BLUEBOOK, OLDCLAIM) and coerce to numeric
```

## Missing Value Imputation

_Describe the imputation strategy and, where relevant, add missingness flag variables._

```{r impute}
#| label: impute

# TODO: missingness flags and imputation
```

## Feature Transformations

_Describe any bucketing, log/sqrt/Box-Cox transforms, or derived variables and the rationale for
each._

```{r transformations}
#| label: transformations

# TODO: bucketing, log/sqrt/Box-Cox transforms, derived variables
```

# Build Models

## Logistic Regression Models (`TARGET_FLAG`)

### Logistic Model 1

_Describe the variable selection rationale for this model._

```{r logit-1}
#| label: logit-1

# TODO: fit first logistic model
```

### Logistic Model 2

```{r logit-2}
#| label: logit-2

# TODO: fit second logistic model
```

### Logistic Model 3

```{r logit-3}
#| label: logit-3

# TODO: fit third logistic model
```

## Linear Regression Models (`TARGET_AMT`)

_Note: `TARGET_AMT` is only meaningfully modeled for crash cases (`TARGET_FLAG == 1`); consider
whether to filter to that subset._

### Linear Model 1

```{r lm-1}
#| label: lm-1

# TODO: fit first linear model
```

### Linear Model 2

```{r lm-2}
#| label: lm-2

# TODO: fit second linear model
```

## Coefficient Discussion

_Discuss whether coefficient signs and magnitudes make sense for both the logistic and linear
models. Flag and discuss any counterintuitive coefficients and whether the variable is retained._

# Select Models

## Multicollinearity

```{r vif}
#| label: vif

# TODO: variance inflation factors for logistic and linear models
```

## Logistic Model Evaluation

```{r logit-eval}
#| label: logit-eval

# TODO: predicted probabilities/classes, confusion matrix, ROC/AUC
```

## Linear Model Evaluation

```{r lm-eval}
#| label: fig-lm-residuals
#| fig-cap: "Residual diagnostics for the selected linear model"

# TODO: residual diagnostic plots
```

## Final Model Selection

_State which logistic model and which linear model are selected as final, and justify the choice
against the criteria discussed above (AIC/log-likelihood/ROC-AUC for the logistic model;
Adjusted R2/RMSE/residual behavior for the linear model; parsimony and interpretability for both)._

# Predictions on Evaluation Data

```{r predictions}
#| label: predictions

# TODO: apply the same preparation steps used on train to evaluation, then predict and write out
```

# Appendix: R Code

_This section is auto-populated by re-echoing the code chunks above when rendering to PDF; no
additional content is needed here unless supplementary code is included._
