---
title: "Homework #1: Moneyball -- Predicting Team Wins"
subtitle: "DATA 621 -- Business Analytics and Data Mining"
author:
  - name: Kevin Havis
  - name: Michael Drake
  - name: Rebecca Bronstein
  - name: Aaliyah John-Harry
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(corrplot)
library(psych)
library(mice)
library(car)
library(caret)
library(VIM)
library(kableExtra)


theme_set(theme_minimal())
```

# Introduction

In this assignment we build a multiple linear regression model to predict `TARGET_WINS`, the
number of wins for a professional baseball team, using team-level batting, baserunning,
pitching, and fielding statistics adjusted to a 162 game season.

# Data Exploration

We first load both data sets. We will drop the index to ensure we do not accidentally use it in our feature set. The goal of data exploration is to inform ourselves of the characteristics of the particular dataset, discover important relationships, and get a general idea of what preprocessing steps may be necessary.

```{r}
#| label: load-data

train <- read_csv("data/moneyball-training-data.csv")
evaluation <- read_csv("data/moneyball-evaluation-data.csv")

train <- train |> select(-INDEX)
```

## Size and Structure

We have a ~2,000 row data set with 16 columns, including our target variable. We immediately notice a high number of missing values in `TEAM_BATTING_HBP` - we will want to investigate this.

```{r}
#| label: data-structure

tibble(
  Variable = names(train),
  Type     = map_chr(train, ~ class(.x)[1]),
  Example  = map_chr(train, ~ paste(head(.x, 3), collapse = ", "))
) |>
  kable()
```

## Summary Statistics

It is often useful to check a summary of descriptive statistics on the data set to get a initial understanding of the data. We're looking for things like high standard deviations, large differences between means and medians (indicating skew), or another anomalous points.

```{r}
#| label: summary-stats

describe(train) |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  select(Variable, mean, sd, median, se) |>
  kable(digits = 2)
```

With this quick glance, we make the below observations; we'll keep this in mind later when we decide what transformations are necessary to appropriately scale the data.

- The variables throughout this dataset have different orders of magnitude
- Most variables are complete (non-null) with the exceptions of `TEAM_BASERUN_SB`, `TEAM_BASERUN SO`,`TEAM_BASERUN_CS`, `TEAM_BATTING_HBP`, and `TEAM_PITCHING_SO`.
- We see a very high degree of skew in `TEAM_PITCHING_H` and `TEAM_PITCHING_SO`
- We see the minimum of most attributes is zero, indicating these should all be positive metrics floored at zero 

<!-- (It seems unrealistic that any values be 0 over a full season. I believe these should be treated as missing values.)
(I also think that some of the maximum values are extremely high and are thus typos. For example, `TEAM_PITCHING_H` and `TEAM_PITCHING_SO` both exceeding 10,000.) -->

## Distributions

Summary statistics will highlight unusual numbers, but visualizing the distributions will allow us to intuit the shape of the variables.

Since these are all continuous quantitative numbers, we visualize all of them as KDE plots.

Our goal here is to identify any potential skewed variables, or otherwise unusual distributions.

```{r}
#| label: fig-distributions
#| fig-cap: "Distributions of predictor and target variables"

# Pivot into long format to make it easier to vizualize
train_long <- pivot_longer(
  train,
  cols = where(is.numeric),   
  names_to = "variable",      
  values_to = "value"         
)

ggplot(train_long, aes(x = value)) +
  geom_density(bins = 20, fill = "#3a86d4", colour = "white") +
  facet_wrap(~ variable, scales = "free") +
  labs(x = NULL, y = "Count", title = "Variable Distributions") +
  theme_minimal(base_size = 9.6)
```

## Boxplots

Boxplots allow us to more easily spot outliers as distinct points, and is a good visual to pair with distribution plots.

```{r}
#| label: fig-boxplots
#| fig-cap: "Boxplots of predictor and target variables"

ggplot(train_long, aes(x = "", y = value)) +
  geom_boxplot(
    fill = "#3a86d4", colour = "black",
    outlier.colour = "black", linewidth = 0.4
  ) +
  facet_wrap(~ variable, scales = "free") +
  labs(x = NULL, y = NULL, title = "Variable Distributions") +
  theme_minimal(base_size = 9.6)
```

## Correlation with Target

Of course, we would like to know what features are correlated with our target variable. A quick look here will help us plan our feature selection.

```{r}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of all variables"
#| fig-width: 12
#| fig-height: 12

# Convert to raw matrix first
cor_matrix <- train |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs")

# Then we can plot the corr plot
corrplot(cor_matrix,
         method = "color",
         type = "lower",
         order = "hclust",
         addCoef.col = "black",
         number.cex = 0.6,
         tl.col = "black",
         tl.cex = 0.6,
         tl.srt = 45,
         diag = FALSE)
```

## Missing Data

Missing data is critical to handle in any machine learning task. We made observations on specific variables earlier, but here we can visualize the proporption of missingness.

```{r}
#| label: fig-missing
#| fig-cap: "Count and proportion of missing values by variable"

missing_props <- train_long |>
  group_by(variable) |>
  summarise(
    proportion_missing = mean(is.na(value)),
    .groups = "drop"
  )

ggplot(missing_props, aes(x = reorder(variable, proportion_missing),
                          y = proportion_missing)) +
  geom_col(fill = "#3a86d4") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion missing", title = "Missing Data by Variable") +
  theme_minimal(base_size = 12)
```


Our exploratory data analysis has revealed some key insights on this dataset that we must address in our modeling, especially to ensure the assumptions of linear regression are met.

- Null values in multiple columns will need to be addressed by imputation or removal
- Variables with high skewness will require mathematical transformations
- Outliers should be handled appropriately, either by removal or transformation
- Highly correlated variables should be considered for removal to avoid multicollinearity issues in the model

# Data Preparation

## Missing Value Imputation

We must now address our strategy for handling missing values.

- `TEAM_BATTING_HBP` is almost always null; any imputation we perform will skew the entire feature to that imputation. We will simply drop it, as it's >90% null.
  - We should address that `TEAM_BATTING_HBP` is significantly null (over 70%) and has a low correlation with the target variable, so we may consider dropping it from the model entirely when we get to feature selection
- `TEAM_BASERUN_CS` has a high degree of right-skew, so we will impute with the median, which is less sensitive to outliers
- `TEAM_BASERUN_SB` is slightly skewed so we will again impute with the median
- `TEAM_PITCHING_SO` has some extreme outliers but is otherwise normally distributed and centered, so we will impute with the mean
- `TEAM_BATTING_SO` has dual modes, so we will use k-nearest neighbors imputation to fill in the missing values, which will take into account the other variables in the dataset.
- `TEAM_FIELDING_DP` has a small number of missing values, so we will impute with the mean given the roughly normal distribution.


```{r}
#| label: impute

train_imputed <- train |>
  mutate(
    TEAM_BASERUN_CS = ifelse(
      is.na(TEAM_BASERUN_CS),
      median(TEAM_BASERUN_CS, na.rm = TRUE),
      TEAM_BASERUN_CS
    ),
    TEAM_BASERUN_SB = ifelse(
      is.na(TEAM_BASERUN_SB),
      median(TEAM_BASERUN_SB, na.rm = TRUE),
      TEAM_BASERUN_SB
    ),
    TEAM_PITCHING_SO = ifelse(
      is.na(TEAM_PITCHING_SO),
      mean(TEAM_PITCHING_SO, na.rm = TRUE),
      TEAM_PITCHING_SO
    ),
    TEAM_FIELDING_DP = ifelse(
      is.na(TEAM_FIELDING_DP),
      mean(TEAM_FIELDING_DP, na.rm = TRUE),
      TEAM_FIELDING_DP
    )
  )

train_imputed <- train_imputed  |>
  VIM::kNN(variable = "TEAM_BATTING_SO", k = 5, imp_var = FALSE)

# drop TEAM_BATTING_HBP since it is almost always null
train_imputed <- train_imputed |>
  select(-TEAM_BATTING_HBP)
```

```{r}
#| label: confirm-missing

# Confirm that there are no more missing values
cat("Remaining missing values:", is.na(train_imputed) |> sum())
```

## Feature Transformations

Now we address any non-linear relationships and/or outliers in our features and our target variable. The easiest way to do this is to visualize the relationship between each predictor and the target variable. A combination of the distribution plots and scatterplots w.r.t the target variable will help us identify which variables may require transformation.

```{r}
#| label: fig-transformations
#| fig-cap: "Scatterplots of predictor variables against target variable"

# Facet wrap scatterplots of each predictor against the target variable
plot_predictors_vs_target <- function(df, title) {
  df |>
    pivot_longer(cols = -TARGET_WINS, names_to = "variable", values_to = "value") |>
    ggplot(aes(x = value, y = TARGET_WINS)) +
    geom_point(alpha = 0.5) +
    facet_wrap(~ variable, scales = "free") +
    labs(x = NULL, y = "TARGET_WINS", title = title) +
    theme_minimal(base_size = 12)
}

plot_predictors_vs_target(
  train_imputed,
  "Scatterplots of Predictor Variables vs Target Variable"
)
```

This gives us a better picture of what variables may required transformation to better fit our linear model. We will break it down by **Outliers**, **Skewness**, and **Non-linear relationships**.

- Outliers: `TEAM_PITCHING_H`, `TEAM_PITCHING_SO`, and `TEAM_PITCHING_BB` have extreme outliers that will need to be addressed. 
- Skewness: Besides outliers, most variables are approximately normally distributed. For now we will leave them as-is to prefer simplicity in our model.
- Non-linear relationships: Our scatterplot does not show any obvious non-linear relationships, but given `TEAM_FIELDING_E` has a long tail, and we see a slight curve in the scatterplot, we may consider a log transformation to see if it improves the linearity of the relationship.


```{r}
#| label: transformations

# Cap values that exceed `threshold` (default: 3 SD above the mean of x itself,
# or pass a threshold learned from a reference set, e.g. a training split)
cap_outliers <- function(x, threshold = mean(x) + 3 * sd(x)) {
  pmin(x, threshold)
}

train_transformed <- train_imputed |>
  mutate(
    # Log transform TEAM_FIELDING_E to reduce skew and improve linearity
    TEAM_FIELDING_E = log(TEAM_FIELDING_E + 1),

    # Cap outliers to three standard deviations from the mean for
    # TEAM_PITCHING_H, TEAM_PITCHING_SO, and TEAM_PITCHING_BB
    TEAM_PITCHING_H  = cap_outliers(TEAM_PITCHING_H),
    TEAM_PITCHING_SO = cap_outliers(TEAM_PITCHING_SO),
    TEAM_PITCHING_BB = cap_outliers(TEAM_PITCHING_BB)
  )
```

## Prepared Data Summary

```{r}
#| label: fig-transformed-scatter
#| fig-cap: "Scatterplots of predictor variables vs target, post-transformation"

plot_predictors_vs_target(
  train_transformed,
  "Scatterplots of Predictor Variables vs Target Variable (Transformed)"
)
```

# Build Models

## Model 1 - Baseline

A simple linear model is always a good starting point. Even if it is not a good model, it gives us a baseline to compare against other models. We will use all predictors in this model, on the entire dataset, double check the as`sumptions of linear regression, and evaluate the model performance.

We should also be mindful of the base rate of wins in the dataset. The mean wins is 80.79 with a $\pm$ 15.7 standard deviation. This means that a naive model that predicts the mean for all teams would have an RMSE of 15.7. Any model we build should be able to beat this baseline.

```{r}
#| label: baseline-stats

# Mean and standard deviation of TARGET_WINS
train_transformed |>
  summarise(mean_wins = mean(TARGET_WINS), sd_wins = sd(TARGET_WINS)) |>
  kable(digits = 2)
```

```{r}
#| label: model-1

lm_all_feat <- lm(TARGET_WINS ~ ., data = train_transformed)
summary(lm_all_feat)
# Plot residuals
par(mfrow = c(2, 2))
plot(lm_all_feat)
```

Let's start with the plot. We see that our residuals vs fitted values is quite close to a horizontal line, which indicates our model is not biased. The Q-Q residuals tells us "how normally distributed are our residuals?", with the diagonal dotted line representing a perfect normal distribution. We see that our residuals are not perfectly normal, but they are close enough to be acceptable (some deviation at the tails is expected for a real world dataset).

Now, reviewing the model summary, we see our adjusted R2 value is 0.3301, which is not an impressive model. This means only 33.01% of the variance in wins is explained by our model. Comparing our RMSE to the baseline, we see our model has an RMSE of 13.04, which is improved over the baseline of 15.7. 

In summary, this basic linear model with all features is not a good model. While it is slightly better than the naive mean model, we should explore further to find a better model.

Before we move on, we should note the p-values for some of the feature coefficients. We see very large p-values for `TEAM_PITCHING_BB`, `TEAM_PITCHING_HR`, and `TEAM_BASERUN_CS`, which indicates these features are not statistically significant in predicting wins. We'll take a look at this next.

## Model 2 - Feature Selection

Let's consider pruning our feature set for our second model. If we review the earlier correlation plot, we see that `TEAM_BATTING_HR` and `TEAM_PITCHING_HR` are very highly correlated at 97%. These features encode the same information, so we will drop them.

Evaluating pair-wise correlations is a good first step, but a more robust way to evaluate multicollinearity is to calculate the variance inflation factor (VIF) for each feature. A VIF value greater than 5 indicates a high degree of multicollinearity, and a value greater than 10 indicates a very high degree of multicollinearity.

We now calculate VIF for our first model to see if any features have a high degree of multicollinearity.

```{r}
#| label: vif-model1

enframe(vif(lm_all_feat), name = "Term", value = "VIF") |>
  kable(digits = 2)
```

As expected, the home run features have a very high VIF, indicating they are highly collinear. We also see that `TEAM_PITCHING_BB` and `TEAM_BATTING_BB` have a similar relationship with a VIF > 5, so we will drop `TEAM_PITCHING_BB`  and `TEAM_PITCHING_HR` from our second model.

```{r}
#| label: model-2-pruning
# Prune features for model 2
train_pruned <- train_transformed |>
  select(-TEAM_PITCHING_BB, -TEAM_PITCHING_HR, -TEAM_BATTING_HR)
```

Let's now fit our second model with the pruned feature set and evaluate the results.

```{r}
#| label: model-2
lm_pruned <- lm(TARGET_WINS ~ ., data = train_pruned)

summary(lm_pruned)
par(mfrow = c(2, 2))
plot(lm_pruned)
```

We do not see a significant improvement in our metrics with this feature selection, but at least our model is simpler with no downsides, having removed the multicollinear features. We will explore one more model with a more robust null value imputation, advanced feature selection, and regularization

## Model 3 - Informed Preprocessing

We will take what we've learned from the data in our last models and apply it here, along with generally more sophisticated techniques. For this third model, we want to apply the below;

- K Nearest Neighbor null imputation
- Backward elimination feature selection

KNN imputation should allow for more representative features rather than mean or median imputation. Backward elmination will ensure we are using the most informative features, while keeping the model as simple as possible. 

```{r}
#| label: model-3-prep

# Impute every remaining missing column with k-nearest neighbors, after
# dropping TEAM_BATTING_HBP (almost always null). Reused later for the
# evaluation data and the holdout validation split.
impute_missing_knn <- function(df, k = 5) {
  df |>
    select(-TEAM_BATTING_HBP) |>
    VIM::kNN(
      variable = c(
        "TEAM_BASERUN_CS", "TEAM_BASERUN_SB", "TEAM_PITCHING_SO",
        "TEAM_BATTING_SO", "TEAM_FIELDING_DP"
      ),
      k = k,
      imp_var = FALSE
    )
}

train_imputed <- train |> impute_missing_knn()

# Same transformations as before
train_transformed <- train_imputed |>
  mutate(
    # Log transform TEAM_FIELDING_E to reduce skew and improve linearity
    TEAM_FIELDING_E = log(TEAM_FIELDING_E + 1),

    # Cap outliers to three standard deviations from the mean for
    # TEAM_PITCHING_H, TEAM_PITCHING_SO, and TEAM_PITCHING_BB
    TEAM_PITCHING_H  = cap_outliers(TEAM_PITCHING_H),
    TEAM_PITCHING_SO = cap_outliers(TEAM_PITCHING_SO),
    TEAM_PITCHING_BB = cap_outliers(TEAM_PITCHING_BB)
  )
```

With the improved preprocessing, we now perform backwards elimination for our feature selection. This means we start with all features, produce a model, remove the most uninformative feature, fit a new model, and so on until the best combination of features is found. This is a simple but effective method of automatic feature selection.

```{r}
#| label: model-3

# Fit the saturated model, then run backward elimination by AIC
lm_full_m3 <- lm(TARGET_WINS ~ ., data = train_transformed)

lm_backward <- step(lm_full_m3, direction = "backward", trace = 1)

summary(lm_backward)

# Plot residuals
par(mfrow = c(2, 2))
plot(lm_backward)
```

Backward elimination only removes `TEAM_PITCHING_HR`, stopping at 13 of the 14 predictors. This agrees with our Model 2 pruning: `TEAM_PITCHING_HR` was the redundant half of its collinear pair with `TEAM_BATTING_HR`, so the algorithm converges on the same conclusion we reached manually with VIF, though it keeps `TEAM_PITCHING_BB` and `TEAM_BATTING_2B`, which we might have been tempted to drop by hand.

Reviewing the metrics, our more robust preprocessing has improved the model materially. Adjusted R2 is 0.3792, versus ~0.31 for Model 1 and Model 2, and RMSE is 12.37, down from 13.04. The residual plots confirm that our residuals are evenly distributed, and all assumptions of linear regresison are still met - expected, as we did not transform the features for this model.

Let's now analyze the model results more deeply.

### Coefficient Discussion

The coefficients should be assessed by two measures; relative magnitude, and directionality. We often do not want to concern ourselves with the specific values of the coefficients, as these are point estimates from the possible values, but the magnitude and direction give us information on our features.

While all of our backwards elimination selected features are statistically significant, we make the below observations regarding some of the more interesting coefficients;

- `TEAM_FIELDING_E` has the largest coefficient by a significant margin, and is negatively correlated with target wins. This tells us that fielding errors is a significant contributor towards losses, which is intuitive and easily explained. A team looking to improve their win rate should first focus on minimizing fielding errors.
- `TEAM_FIELDING_DP` represents double plays, where a fielding team gets two outs in the same play, generally considered very favorable. However, our model has assigned a *negative* correlation, indicating that double plays are associated with losses more than wins.
- `TEAM_BATTING_3B` has the highest *positive* correlation with wins. Interestingly, our model has it as even higher than home runs (`TEAM_BATTING_HR`), although considering difference is within the error ranges, this is a good example of not taking the coefficient values too literally. Batting generaly shows a positive correlation, as expected, with the exception of double bases which is negative.

```{r}
#| label: model-3-coefficients

# Display the coefficients in a clean, readable table
summary(lm_backward)$coefficients |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  rename(`p value` = `Pr(>|t|)`) |>
  kable(digits = 4)
```


### Multicollinearity

Our best model still has some degree of multicolinearity between features as indicated by VIF scores > 5. However, the key thing is that the very tightly coupled, pair-wise features have been removed. As evidenced from our improved R2 values, multicollinear features still contribute meaningful signal in the model, which is why our step feature selection kept them. The very correlated pair features `TEAM_BATTING_HR` and `TEAM_PITCHING_HR` have been excluded from Model 2 and 3 as reasoned previously.

A further improvement would be to apply some form of regularization to the model, such as lasso or ridge regressions. These regressions include a regularization hyperparameter that can be tuned to shrink collinear coeffficients to reduce the impact of the multicolinearity.

```{r}
#| label: vif

list(
  "Model 1: All Features" = lm_all_feat,
  "Model 2: Pruned" = lm_pruned,
  "Model 3: Backward Elimination" = lm_backward
) |>
  map(vif) |>
  map(~ enframe(.x, name = "Term", value = "VIF")) |>
  bind_rows(.id = "Model") |>
  pivot_wider(names_from = Model, values_from = VIF) |>
  kable(digits = 2) |>
  kable_styling(font_size = 8)
```


### Residual Diagnostics

When analyzing residuals, we are concerned with obvious non-linear patterns that indicate homoscedasticity or some non-linearity in the feature-target correlations. Both of these are visual indicators that we could then further investigate.

Throughout all models, the residuals have appeared to be quite reasonable, and as such we can conclude that general linear models are appropriate for this data. The errors are well distributed with no obvious patterns.

```{r}
#| label: fig-residuals
#| fig-cap: "Residual diagnostic plots for the selected model"

par(mfrow = c(2, 2))
plot(lm_backward)
```

## Model Selection

To validate our model selection, we will use the training data and create a holdout set. The bulk of the training data will be used to fit each of Model 1, 2, and 3, and the remaining holdout set will be used to compare out of sample performance. This is an important step, because it confirms not only which model has the most predictive power, but which will generalize to unseen data.

```{r}
#| label: holdout-validation

# Create hold out sets from the training data to compare all three models
set.seed(621)
holdout_index <- createDataPartition(train$TARGET_WINS, p = 0.8, list = FALSE)
train_split <- train[holdout_index, ]
test_split  <- train[-holdout_index, ]

# Model 1 / 2 prep: mean/median + single-variable KNN imputation,
# with imputation statistics learned from train_split only
impute_stats_m1 <- list(
  cs_median = median(train_split$TEAM_BASERUN_CS, na.rm = TRUE),
  sb_median = median(train_split$TEAM_BASERUN_SB, na.rm = TRUE),
  so_mean   = mean(train_split$TEAM_PITCHING_SO, na.rm = TRUE),
  dp_mean   = mean(train_split$TEAM_FIELDING_DP, na.rm = TRUE)
)

impute_mean_median_m1 <- function(df, stats) {
  df |>
    mutate(
      TEAM_BASERUN_CS = ifelse(
        is.na(TEAM_BASERUN_CS), stats$cs_median, TEAM_BASERUN_CS
      ),
      TEAM_BASERUN_SB = ifelse(
        is.na(TEAM_BASERUN_SB), stats$sb_median, TEAM_BASERUN_SB
      ),
      TEAM_PITCHING_SO = ifelse(
        is.na(TEAM_PITCHING_SO), stats$so_mean, TEAM_PITCHING_SO
      ),
      TEAM_FIELDING_DP = ifelse(
        is.na(TEAM_FIELDING_DP), stats$dp_mean, TEAM_FIELDING_DP
      )
    ) |>
    VIM::kNN(variable = "TEAM_BATTING_SO", k = 5, imp_var = FALSE) |>
    select(-TEAM_BATTING_HBP)
}

train_split_imp_m1 <- impute_mean_median_m1(train_split, impute_stats_m1)
test_split_imp_m1  <- impute_mean_median_m1(test_split, impute_stats_m1)

# Cap and transform using thresholds learned from train_split only
cap_stats_m1 <- list(
  ph_thresh = mean(train_split_imp_m1$TEAM_PITCHING_H) +
    3 * sd(train_split_imp_m1$TEAM_PITCHING_H),
  pso_thresh = mean(train_split_imp_m1$TEAM_PITCHING_SO) +
    3 * sd(train_split_imp_m1$TEAM_PITCHING_SO),
  pbb_thresh = mean(train_split_imp_m1$TEAM_PITCHING_BB) +
    3 * sd(train_split_imp_m1$TEAM_PITCHING_BB)
)

prep_transform <- function(df, stats) {
  df |>
    mutate(
      TEAM_FIELDING_E  = log(TEAM_FIELDING_E + 1),
      TEAM_PITCHING_H  = cap_outliers(TEAM_PITCHING_H, stats$ph_thresh),
      TEAM_PITCHING_SO = cap_outliers(TEAM_PITCHING_SO, stats$pso_thresh),
      TEAM_PITCHING_BB = cap_outliers(TEAM_PITCHING_BB, stats$pbb_thresh)
    )
}

train_split_t_m1 <- prep_transform(train_split_imp_m1, cap_stats_m1)
test_split_t_m1  <- prep_transform(test_split_imp_m1, cap_stats_m1)

lm_holdout_m1 <- lm(TARGET_WINS ~ ., data = train_split_t_m1)
lm_holdout_m2 <- lm(
  TARGET_WINS ~ . - TEAM_PITCHING_BB - TEAM_PITCHING_HR - TEAM_BATTING_HR,
  data = train_split_t_m1
)

# Model 3 prep: KNN imputation across all missing columns,
# with capping statistics again learned from train_split only
train_split_imp_m3 <- train_split |> impute_missing_knn()
test_split_imp_m3  <- test_split  |> impute_missing_knn()

cap_stats_m3 <- list(
  ph_thresh = mean(train_split_imp_m3$TEAM_PITCHING_H) +
    3 * sd(train_split_imp_m3$TEAM_PITCHING_H),
  pso_thresh = mean(train_split_imp_m3$TEAM_PITCHING_SO) +
    3 * sd(train_split_imp_m3$TEAM_PITCHING_SO),
  pbb_thresh = mean(train_split_imp_m3$TEAM_PITCHING_BB) +
    3 * sd(train_split_imp_m3$TEAM_PITCHING_BB)
)

train_split_t_m3 <- prep_transform(train_split_imp_m3, cap_stats_m3)
test_split_t_m3  <- prep_transform(test_split_imp_m3, cap_stats_m3)

lm_holdout_m3_full <- lm(TARGET_WINS ~ ., data = train_split_t_m3)
lm_holdout_m3 <- step(lm_holdout_m3_full, direction = "backward", trace = 0)

# Score all three on the held-out split
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
r2 <- function(actual, pred) {
  1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2)
}

pred_m1 <- predict(lm_holdout_m1, newdata = test_split_t_m1)
pred_m2 <- predict(lm_holdout_m2, newdata = test_split_t_m1)
pred_m3 <- predict(lm_holdout_m3, newdata = test_split_t_m3)

# Put results together
tibble(
  Model = c(
    "Model 1: All Features", "Model 2: Pruned", "Model 3: Backward Elimination"
  ),
  `Holdout RMSE` = c(
    rmse(test_split_t_m1$TARGET_WINS, pred_m1),
    rmse(test_split_t_m1$TARGET_WINS, pred_m2),
    rmse(test_split_t_m3$TARGET_WINS, pred_m3)
  ),
  `Holdout R2` = c(
    r2(test_split_t_m1$TARGET_WINS, pred_m1),
    r2(test_split_t_m1$TARGET_WINS, pred_m2),
    r2(test_split_t_m3$TARGET_WINS, pred_m3)
  ),
  `Train RMSE` = c(
    sqrt(mean(resid(lm_holdout_m1)^2)),
    sqrt(mean(resid(lm_holdout_m2)^2)),
    sqrt(mean(resid(lm_holdout_m3)^2))
  ),
  `Train Adj. R2` = c(
    summary(lm_holdout_m1)$adj.r.squared,
    summary(lm_holdout_m2)$adj.r.squared,
    summary(lm_holdout_m3)$adj.r.squared
  )
) |>
  kable(digits = 3)
```

Model 3 wins on both holdout RMSE and holdout R2, and its train/holdout gap is smaller than Model 1 or Model 2's, suggesting it generalizes rather than just overfitting the training data. We select Model 3 as our final model.

### Predictions on Evaluation Data

Now that Model 3 is confirmed as our best model, we run the evaluation dataset through it.

```{r}
#| label: predictions

evaluation_imputed <- evaluation |>
  select(-INDEX) |>
  impute_missing_knn()

# Cap using train's thresholds, not evaluation's own, to avoid leakage
evaluation_transformed <- evaluation_imputed |>
  mutate(
    TEAM_FIELDING_E = log(TEAM_FIELDING_E + 1),
    TEAM_PITCHING_H = cap_outliers(
      TEAM_PITCHING_H,
      mean(train_imputed$TEAM_PITCHING_H) + 3 * sd(train_imputed$TEAM_PITCHING_H)
    ),
    TEAM_PITCHING_SO = cap_outliers(
      TEAM_PITCHING_SO,
      mean(train_imputed$TEAM_PITCHING_SO) + 3 * sd(train_imputed$TEAM_PITCHING_SO)
    ),
    TEAM_PITCHING_BB = cap_outliers(
      TEAM_PITCHING_BB,
      mean(train_imputed$TEAM_PITCHING_BB) + 3 * sd(train_imputed$TEAM_PITCHING_BB)
    )
  )

predictions <- evaluation |>
  transmute(INDEX, TARGET_WINS = predict(lm_backward, newdata = evaluation_transformed))

write_csv(predictions, "data/moneyball-predictions.csv")

head(predictions) |>
  kable()
```

# Conclusion

Starting with data exploration and a note on the base rates set up our expectations and gave us a baseline for our models. Using visualizations, we were able to understand the data more thoroughly which helped inform our preprocessing and feature selection methods. Iterating through multiple models allowed us to find continually improve our model and identify areas of improvement at each step. Evaluating the models comprehensively by considering R2, coefficients, RMSEs, and residual analysis allowed us to be confident our model was appropriate and our process effective.

We selected Model 3, a simple multiple-regression model that used backwards elimination feature selection, as well as KNN imputation as our best option, confirmed by holdout validation against Models 1 and 2. This model is decent compromise of complexity and fit, but we believe further improvements could be made by introducing regularization.

# Appendix: R Code

_This section re-echoes every code chunk above, without re-executing any of it, for reproducibility._

```{r}
#| label: appendix-code
#| ref.label: !expr knitr::all_labels()
#| echo: true
#| eval: false
```
