---
title: "Homework #1: Moneyball -- Predicting Team Wins"
subtitle: "DATA 621 -- Business Analytics and Data Mining"
author:
  - name: Kevin Havis
  - name: Michael Drake
  - name: Rebecca Bronstein
  - name: Aaliyah John-Harry
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(corrplot)
library(psych)
library(mice)
library(car)
library(caret)
library(VIM)
library(kableExtra)


theme_set(theme_minimal())
```

# Introduction

In this assignment we build a multiple linear regression model to predict `TARGET_WINS`, the
number of wins for a professional baseball team, using team-level batting, baserunning,
pitching, and fielding statistics adjusted to a 162 game season.

# Data Exploration

We first load both data sets. We will drop the index to ensure we do not accidentally use it in our feature set. 

```{r load-data}
#| label: load-data

train <- read_csv("hw1/data/moneyball-training-data.csv")
evaluation <- read_csv("hw1/data/moneyball-evaluation-data.csv")

train <- train %>% select(-INDEX)
```

## Size and Structure

We have a ~2,000 row data set with 16 columns, including our target variable. We immediately notice a high number of missing values in `TEAM_BATTING_HBP` - we will want to investigate this.

```{r data-structure}
#| label: data-structure
glimpse(train)
```

## Summary Statistics

It is often useful to check a summary of descriptive statistics on the data set to get a initial understanding of the data. We're looking for things like high standard deviations, large differences between means and medians (indicating skew), or another anomalous points.

```{r summary-stats}
#| label: summary-stats

describe(train)
```

With this quick glance, we make the below observations; we'll keep this in mind later when we decide what transformations are necessary to appropriately scale the data.

- The variables throughout this dataset have different orders of magnitude
- Most variables are complete (non-null) with the exceptions of `TEAM_BASERUN_SB`, `TEAM_BASERUN SO`,`TEAM_BASERUN_CS`, `TEAM_BATTING_HBP`, and `TEAM_PITCHING_SO`.
- We see a very high degree of skew in `TEAM_PITCHING_H` and `TEAM_PITCHING_SO`
- We see the minimum of most attributes is zero, indicating these should all be positive metrics floored at zero 

<!-- (It seems unrealistic that any values be 0 over a full season. I believe these should be treated as missing values.)
(I also think that some of the maximum values are extremely high and are thus typos. For example, `TEAM_PITCHING_H` and `TEAM_PITCHING_SO` both exceeding 10,000.) -->

## Distributions

Summary statistics will highlight unusual numbers, but visualizing the distributions will allow us to intuit the shape of the variables.

Since these are all continuous quantitative numbers, we visualize all of them as KDE plots.

```{r distributions}
#| label: fig-distributions
#| fig-cap: "Distributions of predictor and target variables"

# Pivot into long format to make it easier to vizualize
train_long <- pivot_longer(
  train,
  cols = where(is.numeric),   
  names_to = "variable",      
  values_to = "value"         
)

ggplot(train_long, aes(x = value)) +
  geom_density(bins = 20, fill = "#3a86d4", colour = "white") +
  facet_wrap(~ variable, scales = "free") +
  labs(x = NULL, y = "Count", title='Variable Distributions') +
  theme_minimal(base_size = 12)
```

## Boxplots

Boxplots allow us to more easily spot outliers. 

```{r boxplots}
#| label: fig-boxplots
#| fig-cap: "Boxplots of predictor and target variables"

ggplot(train_long, aes(x = "", y = value)) +
  geom_boxplot(fill = "#3a86d4", colour = "black", outlier.colour = "black", linewidth = 0.4) +
  facet_wrap(~ variable, scales = "free") +
  labs(x = NULL, y = NULL, title = 'Variable Distributions') +
  theme_minimal(base_size = 12)
```

## Correlation with Target

```{r correlation}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of all variables"
#| fig-width: 12
#| fig-height: 12

# Convert to raw matrix first
cor_matrix <- train %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

# Then we can plot the corr plot
corrplot(cor_matrix,
         method = "color",
         type = "lower",
         order = "hclust",
         addCoef.col = "black",
         number.cex = 0.6,
         tl.col = "black",
         tl.cex = 0.6,
         tl.srt = 45,
         diag = FALSE)
```

## Missing Data

Missing data is critical to handle in any machine learning task. We made observations on specific variables earlier, but here we can visualize the proporption of missingness.

```{r missing-data}
#| label: fig-missing
#| fig-cap: "Count and proportion of missing values by variable"

missing_props <- train_long %>%
  group_by(variable) %>%
  summarise(
    proportion_missing = mean(is.na(value)),
    .groups = "drop"
  )

ggplot(missing_props, aes(x = reorder(variable, proportion_missing),
                          y = proportion_missing)) +
  geom_col(fill = "#3a86d4") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion missing", title="Missing Data by Variable") +
  theme_minimal(base_size = 12)
```


Our exploratory data analysis has revealed some key insights on this dataset that we must address in our modeling, especially to ensure the assumptions of linear regression are met.

- Null values in multiple columns will need to be addressed by imputation or removal
- Variables with high skewness will require mathematical transformations
- Outliers should be handled appropriately, either by removal or transformation
- Highly correlated variables should be considered for removal to avoid multicollinearity issues in the model

# Data Preparation

## Missing Value Imputation

We must now address our strategy for handling missing values.

- `TEAM_BATTING_HBP` is almost always null; any imputation we perform will skew the entire feature to that imputation. We will simply drop it, as it's >90% null.
  - We should address that `TEAM_BATTING_HBP` is significantly null (over 70%) and has a low correlation with the target variable, so we may consider dropping it from the model entirely when we get to feature selection
- `TEAM_BASERUN_CS` has a high degree of right-skew, so we will impute with the median, which is less sensitive to outliers
- `TEAM_BASERUN_SB` is slightly skewed so we will again impute with the median
- `TEAM_PITCHING_SO` has some extreme outliers but is otherwise normally distributed and centered, so we will impute with the mean
- `TEAM_BATTING_SO` has dual modes, so we will use k-nearest neighbors imputation to fill in the missing values, which will take into account the other variables in the dataset.
- `TEAM_FIELDING_DP` has a small number of missing values, so we will impute with the mean given the roughly normal distribution.


```{r impute}
#| label: impute

train_imputed <- train |>
  mutate(
    TEAM_BASERUN_CS = ifelse(
      is.na(TEAM_BASERUN_CS),
      median(TEAM_BASERUN_CS, na.rm = TRUE),
      TEAM_BASERUN_CS
    ),
    TEAM_BASERUN_SB = ifelse(
      is.na(TEAM_BASERUN_SB),
      median(TEAM_BASERUN_SB, na.rm = TRUE),
      TEAM_BASERUN_SB
    ),
    TEAM_PITCHING_SO = ifelse(
      is.na(TEAM_PITCHING_SO),
      mean(TEAM_PITCHING_SO, na.rm = TRUE),
      TEAM_PITCHING_SO
    ),
    TEAM_FIELDING_DP = ifelse(
      is.na(TEAM_FIELDING_DP),
      mean(TEAM_FIELDING_DP, na.rm = TRUE),
      TEAM_FIELDING_DP
    )
  )

train_imputed <- train_imputed  |>
  VIM::kNN(variable = "TEAM_BATTING_SO", k = 5, imp_var = FALSE)

# drop TEAM_BATTING_HBP since it is almost always null
train_imputed <- train_imputed |>
  select(-TEAM_BATTING_HBP)
```

```{r}
# Confirm that there are no more missing values
is.na(train_imputed) |> sum()
```

## Feature Transformations

Now we address any non-linear relationships and/or outliers in our features and our target variable. The easiest way to do this is to visualize the relationship between each predictor and the target variable. A combination of the distribution plots and scatterplots w.r.t the target variable will help us identify which variables may require transformation.

```{r}
#| label: fig-transformations
#| fig-cap: "Scatterplots of predictor variables against target variable"

# Facet wrap scatterplots of each predictor against the target variable
train_long_no_target <- train_imputed |>
  pivot_longer(cols = -TARGET_WINS, names_to = "variable", values_to = "value")

ggplot(train_long_no_target, aes(x = value, y = TARGET_WINS)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  labs(x = NULL, y ="TARGET_WINS", title = "Scatterplots of Predictor Variables vs Target Variable") +
  theme_minimal(base_size = 12)

```

This gives us a better picture of what variables may required transformation to better fit our linear model. We will break it down by **Outliers**, **Skewness**, and **Non-linear relationships**.

- Outliers: `TEAM_PITCHING_H`, `TEAM_PITCHING_SO`, and `TEAM_PITCHING_BB` have extreme outliers that will need to be addressed. 
- Skewness: Besides outliers, most variables are approximately normally distributed. For now we will leave them as-is to prefer simplicity in our model.
- Non-linear relationships: Our scatterplot does not show any obvious non-linear relationships, but given `TEAM_FIELDING_E` has a long tail, and we see a slight curve in the scatterplot, we may consider a log transformation to see if it improves the linearity of the relationship.


```{r transformations}
#| label: transformations
train_transformed <- train_imputed  |> 
  mutate(
    # Log transform TEAM_FIELDING_E to reduce skew and improve linearity
    TEAM_FIELDING_E = log(TEAM_FIELDING_E + 1),

    # Cap outliers to three standard deviations from the mean for
    # TEAM_PITCHING_H, TEAM_PITCHING_SO, and TEAM_PITCHING_BB
    TEAM_PITCHING_H = ifelse(
      TEAM_PITCHING_H > mean(TEAM_PITCHING_H) + 3 * sd(TEAM_PITCHING_H),
      mean(TEAM_PITCHING_H) + 3 * sd(TEAM_PITCHING_H),
      TEAM_PITCHING_H
    ),
    TEAM_PITCHING_SO = ifelse(
      TEAM_PITCHING_SO > mean(TEAM_PITCHING_SO) + 3 * sd(TEAM_PITCHING_SO),
      mean(TEAM_PITCHING_SO) + 3 * sd(TEAM_PITCHING_SO),
      TEAM_PITCHING_SO
    ),
    TEAM_PITCHING_BB = ifelse(
      TEAM_PITCHING_BB > mean(TEAM_PITCHING_BB) + 3 * sd(TEAM_PITCHING_BB),
      mean(TEAM_PITCHING_BB) + 3 * sd(TEAM_PITCHING_BB),
      TEAM_PITCHING_BB
    )
  )

```

## Prepared Data Summary

```{r prepared-summary}
#| label: prepared-summary

head(train_transformed)  |> kable() |> kable_styling(full_width = FALSE, position = "center")

```

```{r}
# Facet wrap scatterplots of each predictor against the target variable
train_transform_long_no_target <- train_transform |>
  pivot_longer(cols = -TARGET_WINS, names_to = "variable", values_to = "value")

ggplot(train_transform_long_no_target, aes(x = value, y = TARGET_WINS)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  labs(x = NULL, y ="TARGET_WINS", title = "Scatterplots of Predictor Variables vs Target Variable") +
  theme_minimal(base_size = 12)
```

# Build Models

## Model 1

A simple linear model is always a good starting point. Even if it is not a good model, it gives us a baseline to compare against other models. We will use all predictors in this model, on the entire dataset, double check the as`sumptions of linear regression, and evaluate the model performance.

We should also be mindful of the base rate of wins in the dataset. The mean wins is 80.79 with a $\pm$ 15.7 standard deviation. This means that a naive model that predicts the mean for all teams would have an RMSE of 15.7. Any model we build should be able to beat this baseline.

```{r}
#| label: baseline-stats
# Mean and standard deviation of TARGET_WINS
train_transformed %>% summarise(mean_wins = mean(TARGET_WINS), sd_wins = sd(TARGET_WINS))

```

```{r model-1}
#| label: all-features-model

lm_all_feat <- lm(TARGET_WINS ~ ., data = train_transformed)
summary(lm_all_feat)
# Plot residuals
par(mfrow = c(2, 2))
plot(lm_all_feat)
```

Let's start with the plot. We see that our residuals vs fitted values is quite close to a horizontal line, which indicates our model is not biased. The Q-Q residuals tells us "how normally distributed are our residuals?", with the diagonal dotted line representing a perfect normal distribution. We see that our residuals are not perfectly normal, but they are close enough to be acceptable (some deviation at the tails is expected for a real world dataset).

Now, reviewing the model summary, we see our adjusted R2 value is 0.3146, which is not an impressive model. This means only 31.46% of the variance in wins is explained by our model. Comparing our RMSE to the baseline, we see our model has an RMSE of 13.04, which is improved over the baseline of 15.7. 

In summary, this basic linear model with all features is not a good model. While it is slightly better than the naive mean model, we should explore further to find a better model.

Before we move on, we should note the p-values for some of the feature coefficients. We see very large p-values for `TEAM_PITCHING_BB`, `TEAM_PITCHING_HR`, and `TEAM_BASERUN_CS`, which indicates these features are not statistically significant in predicting wins. We'll take a look at this next.

## Model 2

Let's consider pruning our feature set for our second model. If we review the earlier correlation plot, we see that `TEAM_BATTING_HR` and `TEAM_PITCHING_HR` are very highly correlated at 97%. These features encode the same information, so we will drop them.

Evaluating pair-wise correlations is a good first step, but a more robust way to evaluate multicollinearity is to calculate the variance inflation factor (VIF) for each feature. A VIF value greater than 5 indicates a high degree of multicollinearity, and a value greater than 10 indicates a very high degree of multicollinearity.

We now calculate VIF for our first model to see if any features have a high degree of multicollinearity.

```{r vif-model1}
#| label: vif-model1
vif(lm_all_feat)
```

As expected, the home run features have a very high VIF, indicating they are highly collinear. We also see that `TEAM_PITCHING_BB` and `TEAM_BATTING_BB` have a similar relationship with a VIF > 5, so we will drop `TEAM_PITCHING_BB`  and `TEAM_PITCHING_HR` from our second model.

```{r}
#| label: model-2-pruning
# Prune features for model 2
train_pruned <- train_transformed |>
  select(-TEAM_PITCHING_BB, -TEAM_PITCHING_HR, -TEAM_BATTING_HR)
```

Let's now fit our second model with the pruned feature set and evaluate the results.

```{r model-2}
#| label: model-2
lm_pruned <- lm(TARGET_WINS ~ ., data = train_pruned)

summary(lm_pruned)
plot(lm_pruned)

```

## Model 3

_Describe the variable selection rationale for this model._

```{r model-3}
#| label: model-3

# TODO: fit third model
```

## Coefficient Discussion

_Discuss whether coefficient signs and magnitudes align with the theoretical effects noted in the
assignment (e.g., positive vs. negative impact on wins). Flag and discuss any counterintuitive
coefficients and whether the variable is retained._

# Select Models

## Multicollinearity

```{r vif}
#| label: vif

# TODO: variance inflation factors
```

## Model Comparison

```{r model-comparison}
#| label: model-comparison

# TODO: compare models on Adjusted R2, RMSE, F-statistic, AIC/BIC
```

## Residual Diagnostics

```{r residual-diagnostics}
#| label: fig-residuals
#| fig-cap: "Residual diagnostic plots for the selected model"

# TODO: residual diagnostic plots
```

## Final Model Selection

_State which model is selected as final and justify the choice against the criteria discussed
above (Adjusted R2, RMSE, F-statistic, parsimony, interpretability, residual behavior)._

# Predictions on Evaluation Data

```{r predictions}
#| label: predictions

# TODO: apply the same preparation steps used on train to evaluation, then predict and write out
```

# Appendix: R Code

_This section is auto-populated by re-echoing the code chunks above when rendering to PDF; no
additional content is needed here unless supplementary code is included._
