---
title: "Homework #1: Moneyball -- Predicting Team Wins"
subtitle: "DATA 621 -- Business Analytics and Data Mining"
author:
  - name: Kevin Havis
  - name: Michael Drake
  - name: Rebecca Bronsteain
  - name: Aaliyah John-Henry
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(corrplot)
library(psych)
library(mice)
library(car)
library(caret)

theme_set(theme_minimal())
```

# Introduction

In this assignment we build a multiple linear regression model to predict `TARGET_WINS`, the
number of wins for a professional baseball team, using team-level batting, baserunning,
pitching, and fielding statistics adjusted to a 162 game season.

# Data Exploration

```{r load-data}
#| label: load-data

train <- read_csv("data/moneyball-training-data.csv")
evaluation <- read_csv("data/moneyball-evaluation-data.csv")

train <- train %>% select(-INDEX)
```

## Size and Structure

```{r data-structure}
#| label: data-structure

# TODO: dim(), glimpse(), etc.
```

## Summary Statistics

```{r summary-stats}
#| label: summary-stats

# TODO: mean / sd / median summary table
```

## Distributions

```{r distributions}
#| label: fig-distributions
#| fig-cap: "Distributions of predictor and target variables"

# TODO: histograms or density plots of each variable
```

## Boxplots

```{r boxplots}
#| label: fig-boxplots
#| fig-cap: "Boxplots of predictor and target variables"

# TODO: boxplots of each variable
```

## Correlation with Target

```{r correlation}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of all variables"

# TODO: correlation matrix / correlation with TARGET_WINS
```

## Missing Data

```{r missing-data}
#| label: fig-missing
#| fig-cap: "Count and proportion of missing values by variable"

# TODO: count / proportion of missing values per variable
```

_Discuss what the exploration reveals: notable distributions, skew, outliers, correlations with
`TARGET_WINS`, multicollinearity among predictors, and which variables have meaningful amounts of
missing data._

# Data Preparation

## Missing Value Imputation

_Describe the imputation strategy (e.g., median imputation, or multiple imputation via `mice`)
and, where relevant, add missingness flag variables._

```{r missing-flags}
#| label: missing-flags

# TODO: create missingness flag variables
```

```{r impute}
#| label: impute

# TODO: impute missing values
```

## Feature Transformations

_Describe any bucketing, log/sqrt/Box-Cox transforms, or derived ratio/combination variables and
the rationale for each._

```{r transformations}
#| label: transformations

# TODO: bucketing, log/sqrt/Box-Cox transforms, derived/combined variables
```

## Prepared Data Summary

```{r prepared-summary}
#| label: prepared-summary

# TODO: summarize the prepared data set
```

# Build Models

## Model 1

_Describe the variable selection rationale for this model._

```{r model-1}
#| label: model-1

# TODO: fit first model
```

## Model 2

_Describe the variable selection rationale for this model (e.g., transformed variables, subset of
predictors)._

```{r model-2}
#| label: model-2

# TODO: fit second model
```

## Model 3

_Describe the variable selection rationale for this model._

```{r model-3}
#| label: model-3

# TODO: fit third model
```

## Coefficient Discussion

_Discuss whether coefficient signs and magnitudes align with the theoretical effects noted in the
assignment (e.g., positive vs. negative impact on wins). Flag and discuss any counterintuitive
coefficients and whether the variable is retained._

# Select Models

## Multicollinearity

```{r vif}
#| label: vif

# TODO: variance inflation factors
```

## Model Comparison

```{r model-comparison}
#| label: model-comparison

# TODO: compare models on Adjusted R2, RMSE, F-statistic, AIC/BIC
```

## Residual Diagnostics

```{r residual-diagnostics}
#| label: fig-residuals
#| fig-cap: "Residual diagnostic plots for the selected model"

# TODO: residual diagnostic plots
```

## Final Model Selection

_State which model is selected as final and justify the choice against the criteria discussed
above (Adjusted R2, RMSE, F-statistic, parsimony, interpretability, residual behavior)._

# Predictions on Evaluation Data

```{r predictions}
#| label: predictions

# TODO: apply the same preparation steps used on train to evaluation, then predict and write out
```

# Appendix: R Code

_This section is auto-populated by re-echoing the code chunks above when rendering to PDF; no
additional content is needed here unless supplementary code is included._
