---
title: "Homework #1: Moneyball -- Predicting Team Wins"
subtitle: "DATA 621 -- Business Analytics and Data Mining"
author:
  - name: Kevin Havis
  - name: Michael Drake
  - name: Rebecca Bronsteain
  - name: Aaliyah John-Henry
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(corrplot)
library(psych)
library(mice)
library(car)
library(caret)


theme_set(theme_minimal())
```

# Introduction

In this assignment we build a multiple linear regression model to predict `TARGET_WINS`, the
number of wins for a professional baseball team, using team-level batting, baserunning,
pitching, and fielding statistics adjusted to a 162 game season.

# Data Exploration

We first load both data sets. We will drop the index to ensure we do not accidentally use it in our feature set. 

```{r load-data}
#| label: load-data

train <- read_csv("hw1/data/moneyball-training-data.csv")
evaluation <- read_csv("hw1/data/moneyball-evaluation-data.csv")

train <- train %>% select(-INDEX)
```

## Size and Structure

We have a ~2,000 row data set with 16 columns, including our target variable. We immediately notice a high number of missing values in `TEAM_BATTING_HBP` - we wil want to investigate this.

```{r data-structure}
#| label: data-structure

glimpse(train)
```

## Summary Statistics

It is often useful to check a summary of descriptive statistics on the data set to get a initial understanding of the data. We're looking for things like high standard deviations, large differences between means and medians (indicating skew), or another anomalous points.

```{r summary-stats}
#| label: summary-stats

describe(train)
```

With this quick glance, we make the below observations; we'll keep this in mind later when we decide what transformations are necessary to appropriately scale the data.

- The variables throughout this dataset have different orders of magnitude
- Most variables are complete (non-null) with the exceptions of `TEAM_BASERUN_SB`, `TEAM_BASERUN SO`,`TEAM_BASERUN_CS`, `TEAM_BATTING_HBP`, and `TEAM_PITCHING_SO`.
- We see a very high degree of skew in `TEAM_PITCHING_H` and `TEAM_PITCHING_SO`
- We see the minimum of most attributes is zero, indicating these should all be positive metrics floored at zero

## Distributions

Summary statistics will highlight unusual numbers, but visualizing the distributions will allow us to intuit the shape of the variables.

Since these are all continuous quantitative numbers, we visualize all of them as KDE plots.

```{r distributions}
#| label: fig-distributions
#| fig-cap: "Distributions of predictor and target variables"

# Pivot into long format to make it easier to vizualize
train_long <- pivot_longer(
  train,
  cols = where(is.numeric),   
  names_to = "variable",      
  values_to = "value"         
)

ggplot(train_long, aes(x = value)) +
  geom_density(bins = 20, fill = "#3a86d4", colour = "white") +
  facet_wrap(~ variable, scales = "free") +
  labs(x = NULL, y = "Count", title='Variable Distributions') +
  theme_minimal(base_size = 12)
```

## Boxplots

Boxplots allow us to more easily spot outliers. 

```{r boxplots}
#| label: fig-boxplots
#| fig-cap: "Boxplots of predictor and target variables"

ggplot(train_long, aes(x = "", y = value)) +
  geom_boxplot(fill = "#3a86d4", colour = "black", outlier.colour = "black", linewidth = 0.4) +
  facet_wrap(~ variable, scales = "free") +
  labs(x = NULL, y = NULL, title = 'Variable Distributions') +
  theme_minimal(base_size = 12)
```

## Correlation with Target

```{r correlation}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of all variables"
#| fig-width: 12
#| fig-height: 12

# Convert to raw matrix first
cor_matrix <- train %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

# Then we can plot the corr plot
corrplot(cor_matrix,
         method = "color",
         type = "lower",
         order = "hclust",
         addCoef.col = "black",
         number.cex = 0.6,
         tl.col = "black",
         tl.cex = 0.6,
         tl.srt = 45,
         diag = FALSE)
```

## Missing Data

Missing data is critical to handle in any machine learning task. We made observations on specific variables earlier, but here we can visualize the proporption of missingness.

```{r missing-data}
#| label: fig-missing
#| fig-cap: "Count and proportion of missing values by variable"

missing_props <- train_long %>%
  group_by(variable) %>%
  summarise(
    proportion_missing = mean(is.na(value)),
    .groups = "drop"
  )

ggplot(missing_props, aes(x = reorder(variable, proportion_missing),
                          y = proportion_missing)) +
  geom_col(fill = "#3a86d4") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion missing", title="Missing Data by Variable") +
  theme_minimal(base_size = 12)
```

_Discuss what the exploration reveals: notable distributions, skew, outliers, correlations with
`TARGET_WINS`, multicollinearity among predictors, and which variables have meaningful amounts of
missing data._

Our exploratory data analysis has revealed some key insights on this dataset that we must address in our modeling, especially to ensure the assumptions of linear regression are met.

- Null values in multiple columns will need to be addressed by imputation or removal
- Variables with high skewness will require mathematical transformations
- Outliers should be handled appropriately, either by removal or transformation
- Highly correlated variables should be considered for removal to avoid multicollinearity issues in the model

# Data Preparation

## Missing Value Imputation

_Describe the imputation strategy (e.g., median imputation, or multiple imputation via `mice`)
and, where relevant, add missingness flag variables._

```{r missing-flags}
#| label: missing-flags

# TODO: create missingness flag variables
```

```{r impute}
#| label: impute

# TODO: impute missing values
```

## Feature Transformations

_Describe any bucketing, log/sqrt/Box-Cox transforms, or derived ratio/combination variables and
the rationale for each._

```{r transformations}
#| label: transformations

# TODO: bucketing, log/sqrt/Box-Cox transforms, derived/combined variables
```

## Prepared Data Summary

```{r prepared-summary}
#| label: prepared-summary

# TODO: summarize the prepared data set
```

# Build Models

## Model 1

_Describe the variable selection rationale for this model._

```{r model-1}
#| label: model-1

# TODO: fit first model
```

## Model 2

_Describe the variable selection rationale for this model (e.g., transformed variables, subset of
predictors)._

```{r model-2}
#| label: model-2

# TODO: fit second model
```

## Model 3

_Describe the variable selection rationale for this model._

```{r model-3}
#| label: model-3

# TODO: fit third model
```

## Coefficient Discussion

_Discuss whether coefficient signs and magnitudes align with the theoretical effects noted in the
assignment (e.g., positive vs. negative impact on wins). Flag and discuss any counterintuitive
coefficients and whether the variable is retained._

# Select Models

## Multicollinearity

```{r vif}
#| label: vif

# TODO: variance inflation factors
```

## Model Comparison

```{r model-comparison}
#| label: model-comparison

# TODO: compare models on Adjusted R2, RMSE, F-statistic, AIC/BIC
```

## Residual Diagnostics

```{r residual-diagnostics}
#| label: fig-residuals
#| fig-cap: "Residual diagnostic plots for the selected model"

# TODO: residual diagnostic plots
```

## Final Model Selection

_State which model is selected as final and justify the choice against the criteria discussed
above (Adjusted R2, RMSE, F-statistic, parsimony, interpretability, residual behavior)._

# Predictions on Evaluation Data

```{r predictions}
#| label: predictions

# TODO: apply the same preparation steps used on train to evaluation, then predict and write out
```

# Appendix: R Code

_This section is auto-populated by re-echoing the code chunks above when rendering to PDF; no
additional content is needed here unless supplementary code is included._
