---
title: "Homework #2: Classification Metrics"
subtitle: "DATA 621 -- Business Analytics and Data Mining"
author:
  - name: Kevin Havis
  - name: Michael Drake
  - name: Rebecca Bronsteain
  - name: Aaliyah John-Henry
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(caret)
library(pROC)

theme_set(theme_minimal())
```

# Introduction

In this assignment we implement, from first principles, a set of classification metrics used to
evaluate a binary classifier, then compare our implementations against the `caret` and `pROC`
packages.

# Data

```{r load-data}
#| label: load-data

scored_data <- read_csv("data/classification-output-data.csv")

glimpse(scored_data)
```

The columns of interest are:

- `class`: the actual class for the observation
- `scored.class`: the predicted class for the observation (threshold of 0.5)
- `scored.probability`: the predicted probability of success for the observation

# Confusion Matrix

```{r confusion-matrix}
#| label: confusion-matrix

# TODO: use table() to produce the raw confusion matrix
```

_Discuss what the rows and columns represent (predicted vs. actual) and confirm the orientation
before proceeding._

# Metric Functions

```{r metric-helpers}
#| label: metric-helpers

# TODO: shared helper to extract TP/FP/TN/FN counts from a scored data frame, if desired
```

## Accuracy

```{r accuracy}
#| label: accuracy

# TODO: function(df, actual_col, predicted_col) -> accuracy
```

## Classification Error Rate

```{r class-error-rate}
#| label: class-error-rate

# TODO: function(df, actual_col, predicted_col) -> classification error rate
# Sanity check: accuracy + error rate should equal 1
```

## Precision

```{r precision}
#| label: precision

# TODO: function(df, actual_col, predicted_col) -> precision
```

## Sensitivity (Recall)

```{r sensitivity}
#| label: sensitivity

# TODO: function(df, actual_col, predicted_col) -> sensitivity
```

## Specificity

```{r specificity}
#| label: specificity

# TODO: function(df, actual_col, predicted_col) -> specificity
```

## F1 Score

```{r f1-score}
#| label: f1-score

# TODO: function(df, actual_col, predicted_col) -> F1 score
```

### Bounds of the F1 Score

_Show algebraically that the F1 score is always bounded between 0 and 1, using the hint that if
$0 < a < 1$ and $0 < b < 1$ then $ab < a$._

# ROC Curve and AUC

```{r roc-function}
#| label: fig-roc-curve
#| fig-cap: "ROC curve computed from scratch"

# TODO: function(df, actual_col, prob_col) -> list(plot = ..., auc = ...)
# Recommended: threshold sequence from 0 to 1 by 0.01, trapezoidal rule for AUC
```

# All Metrics on the Provided Data Set

```{r all-metrics}
#| label: all-metrics

# TODO: apply all functions above to scored_data and present results
```

# Comparison with `caret`

```{r caret-comparison}
#| label: caret-comparison

# TODO: caret::confusionMatrix, sensitivity, specificity
```

_Discuss how the `caret` results compare with the hand-written functions._

# Comparison with `pROC`

```{r proc-comparison}
#| label: fig-proc-roc
#| fig-cap: "ROC curve generated by pROC"

# TODO: pROC::roc / pROC::auc
```

_Discuss how the `pROC` results compare with the hand-written ROC/AUC function._

# Appendix: R Code

_This section is auto-populated by re-echoing the code chunks above when rendering to PDF; no
additional content is needed here unless supplementary code is included._
