---
title: "Homework #3: Logistic Regression -- Predicting Neighborhood Crime Risk"
subtitle: "DATA 621 -- Business Analytics and Data Mining"
author:
  - name: Kevin Havis
  - name: Michael Drake
  - name: Rebecca Bronsteain
  - name: Aaliyah John-Henry
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(corrplot)
library(psych)
library(car)
library(caret)
library(pROC)

theme_set(theme_minimal())
```

# Introduction

In this assignment we build a binary logistic regression model to predict `target`, whether a
neighborhood's crime rate is above the median (1) or not (0), using demographic, housing, and
environmental predictors.

# Data Exploration

```{r load-data}
#| label: load-data

train <- read_csv("data/crime-training-data_modified.csv")
evaluation <- read_csv("data/crime-evaluation-data_modified.csv")
```

## Size and Structure

```{r data-structure}
#| label: data-structure

# TODO: dim(), glimpse(), etc.
```

## Summary Statistics

```{r summary-stats}
#| label: summary-stats

# TODO: mean / sd / median summary table
```

## Distributions

```{r distributions}
#| label: fig-distributions
#| fig-cap: "Distributions of predictor variables"

# TODO: histograms or density plots of each variable
```

## Boxplots by Target

```{r boxplots}
#| label: fig-boxplots
#| fig-cap: "Boxplots of predictors split by target class"

# TODO: boxplots of each predictor split by target
```

## Correlation with Target

```{r correlation}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of all variables"

# TODO: correlation matrix / correlation with target
```

## Missing Data

```{r missing-data}
#| label: missing-data

# TODO: count of missing values per variable
```

_Discuss what the exploration reveals: notable distributions, skew, outliers, class balance of
`target`, correlations with the target, and multicollinearity among predictors._

# Data Preparation

## Missing Value Imputation

_Describe the imputation strategy and, where relevant, add missingness flag variables._

```{r impute}
#| label: impute

# TODO: missingness flags and imputation
```

## Feature Transformations

_Describe any bucketing, log/sqrt/Box-Cox transforms, or derived variables and the rationale for
each._

```{r transformations}
#| label: transformations

# TODO: bucketing, log/sqrt/Box-Cox transforms, derived variables
```

# Build Models

## Model 1

_Describe the variable selection rationale for this model._

```{r model-1}
#| label: model-1

# TODO: fit first logistic model
```

## Model 2

_Describe the variable selection rationale for this model (e.g., stepwise selection, transformed
variables)._

```{r model-2}
#| label: model-2

# TODO: fit second logistic model
```

## Model 3

_Describe the variable selection rationale for this model._

```{r model-3}
#| label: model-3

# TODO: fit third logistic model
```

## Coefficient Discussion

_Discuss whether coefficient signs and magnitudes make sense (e.g., odds ratios via
`exp(coef(model))`). Flag and discuss any counterintuitive coefficients and whether the variable
is retained._

# Select Models

## Multicollinearity

```{r vif}
#| label: vif

# TODO: variance inflation factors
```

## Model Comparison

```{r model-comparison}
#| label: model-comparison

# TODO: compare models on AIC, log-likelihood, McFadden's pseudo-R2
```

## Classification Performance

```{r classification-metrics}
#| label: classification-metrics

# TODO: predicted probabilities/classes, confusion matrix
```

```{r roc-auc}
#| label: fig-roc
#| fig-cap: "ROC curve for the selected model"

# TODO: ROC curve and AUC
```

## Final Model Selection

_State which model is selected as final and justify the choice against the criteria discussed
above (AIC, log-likelihood, ROC/AUC, parsimony, interpretability)._

# Predictions on Evaluation Data

```{r predictions}
#| label: predictions

# TODO: apply the same preparation steps used on train to evaluation, then predict and write out
```

# Appendix: R Code

_This section is auto-populated by re-echoing the code chunks above when rendering to PDF; no
additional content is needed here unless supplementary code is included._
