---
title: "Homework #5: Count Regression -- Predicting Wine Case Sales"
subtitle: "DATA 621 -- Business Analytics and Data Mining"
author:
  - name: Kevin Havis
  - name: Michael Drake
  - name: Rebecca Bronsteain
  - name: Aaliyah John-Henry
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(corrplot)
library(psych)
library(MASS)
library(pscl)
library(car)

theme_set(theme_minimal())
```

# Introduction

In this assignment we build count regression models to predict `TARGET`, the number of sample
cases of wine purchased by distributors, using the wine's chemical properties along with expert
ratings and label appeal.

# Data Exploration

```{r load-data}
#| label: load-data

train <- read_csv("data/wine-training-data.csv")
evaluation <- read_csv("data/wine-evaluation-data.csv")

train <- train %>% select(-INDEX)
```

## Size and Structure

```{r data-structure}
#| label: data-structure

# TODO: dim(), glimpse(), etc.
```

## Summary Statistics

```{r summary-stats}
#| label: summary-stats

# TODO: mean / sd / median summary table
```

## Distribution of the Target

```{r target-distribution}
#| label: fig-target-distribution
#| fig-cap: "Distribution of TARGET (cases purchased)"

# TODO: bar chart of TARGET; compare mean vs. variance to check Poisson dispersion assumption
```

## Distributions of Predictors

```{r distributions}
#| label: fig-distributions
#| fig-cap: "Distributions of predictor variables"

# TODO: histograms or density plots of predictor variables
```

## Correlation with Target

```{r correlation}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of all variables"

# TODO: correlation matrix / correlation with TARGET
```

## Missing Data

```{r missing-data}
#| label: fig-missing
#| fig-cap: "Proportion of missing values by variable"

# TODO: count / proportion of missing values per variable
```

_Discuss what the exploration reveals: the shape and dispersion of `TARGET` relative to Poisson
assumptions, notable predictor distributions, correlations with the target, and which variables
have meaningful missingness (recall the hint that missingness itself, e.g. of `STARS`, may be
predictive)._

# Data Preparation

## Missingness Flags and Imputation

_Describe the imputation strategy. Given the assignment's hint, consider retaining a missingness
flag for `STARS` (and any other variable) as a predictor in its own right rather than only
imputing it away._

```{r missing-flags-impute}
#| label: missing-flags-impute

# TODO: missingness flags and imputation
```

## Feature Transformations

_Describe any bucketing, log/sqrt/Box-Cox transforms, or derived variables and the rationale for
each._

```{r transformations}
#| label: transformations

# TODO: bucketing, log/sqrt/Box-Cox transforms, derived variables
```

# Build Models

## Poisson Regression Models

### Poisson Model 1

_Describe the variable selection rationale for this model._

```{r poisson-1}
#| label: poisson-1

# TODO: fit first Poisson model
```

### Poisson Model 2

```{r poisson-2}
#| label: poisson-2

# TODO: fit second Poisson model
```

## Negative Binomial Regression Models

### Negative Binomial Model 1

```{r nb-1}
#| label: nb-1

# TODO: fit first negative binomial model
```

### Negative Binomial Model 2

```{r nb-2}
#| label: nb-2

# TODO: fit second negative binomial model
```

## (Optional) Zero-Inflated Models

```{r zero-inflated}
#| label: zero-inflated

# TODO: zero-inflated Poisson / negative binomial models, if pursued
```

## Multiple Linear Regression Models

### Linear Model 1

```{r lm-1}
#| label: lm-1

# TODO: fit first linear model
```

### Linear Model 2

```{r lm-2}
#| label: lm-2

# TODO: fit second linear model
```

## Coefficient Discussion

_Discuss whether coefficient signs and magnitudes make sense across model types (Poisson vs.
negative binomial vs. linear), noting cases where the sign or magnitude of a predictor (e.g. `pH`)
differs across models, per the assignment's example. Flag and discuss any counterintuitive
coefficients and whether the variable is retained._

# Select Models

## Dispersion Check

```{r dispersion-check}
#| label: dispersion-check

# TODO: likelihood ratio test comparing Poisson vs. negative binomial fit
```

## Multicollinearity

```{r vif}
#| label: vif

# TODO: variance inflation factors
```

## Model Comparison

```{r model-comparison}
#| label: model-comparison

# TODO: compare models on AIC, average squared error
```

## Final Model Selection

_State which count regression model is selected as final and justify the choice against the
criteria discussed above (AIC, average squared error, parsimony, interpretability). Note that a
count regression model must be selected for deployment even if a linear model performs
comparably or better._

# Predictions on Evaluation Data

```{r predictions}
#| label: predictions

# TODO: apply the same preparation steps used on train to evaluation, then predict and write out
```

# Appendix: R Code

_This section is auto-populated by re-echoing the code chunks above when rendering to PDF; no
additional content is needed here unless supplementary code is included._
